---
title: "Assignment 5: Logistic Regression"
editor_options: 
  chunk_output_type: console
---

## Setup 

Load the following packages: `readr`, `tibble`, `dplyr`, `tidyr`, `ggplot2`, and `mfx`.

## Problem 1. 

Create the `data_analysis` data frame you were using in assignment 4. If the data frame is still loaded in your R session, then you can skip this. 

```{r}
# Write your code here
```

## Problem 2. 

Fit a multiple logistic regression model using `Q222_VOTE_NATIONAL` as the outcome and `AGE_STD`, `Q260_SEX`, `Q287_SES`, `Q288_INCOME`, `Q216_ENCOURAGE_VOTE`, and `Q252_SAT_POL_SYSTEM` as the predictors. Name this model `mod_full`.

```{r}
# Write code here

```

## Problem 3. 

Using the results of `mod_full` interpret the effect of `AGE_STD`. Note if the effect is significant and explain how you came to that determination. 

Intrepret the coefficient by transforming the coefficient into an odds ratio. Then create a plot of the predicted probability of voting as `AGE_STD` changes. Finally, use `mfx::logitmfx` to understand how much the probability of voting changes as `AGE_STD` changes, on average. 

```{r}
# Write code here
```

### Write your interpretation below here

## Problem 4. 

Using the results of `mod_full` interpret the effect of `Q260_SEX`. Note if the effect is significant and explain how you came to that determination. 

Intrepret the coefficient by transforming the coefficient into an odds ratio. Then create a plot of the predicted probability of voting as `Q260_SEX` changes. Finally, use `mfx::logitmfx` to understand how much the probability of voting changes as `Q260_SEX` changes, on average. 

```{r}
# Write code here
```

### Write your interpretation below here

## Problem 5. 

Using the results of `mod_full` interpret the effect of `Q287_SES`. Note if the effect is significant and explain how you came to that determination. 

Intrepret the coefficient by transforming the coefficient into an odds ratio. Then create a plot of the predicted probability of voting as `Q287_SES` changes. Finally, use `mfx::logitmfx` to understand how much the probability of voting changes as `Q287_SES` changes, on average. 

```{r}
# Write code here
```

### Write your interpretation below here

## Problem 6. 

Using the results of `mod_full` interpret the effect of `Q288_INCOME`. Note if the effect is significant and explain how you came to that determination. 

Intrepret the coefficient by transforming the coefficient into an odds ratio. Then create a plot of the predicted probability of voting as `Q288_INCOME` changes. Finally, use `mfx::logitmfx` to understand how much the probability of voting changes as `Q288_INCOME` changes, on average. 

```{r}
# Write code here
```

### Write your interpretation below here

## Problem 7. 

Using the results of `mod_full` interpret the effect of `Q216_ENCOURAGE_VOTE`. Note if the effect is significant and explain how you came to that determination. 

Intrepret the coefficient by transforming the coefficient into an odds ratio. Then create a plot of the predicted probability of voting as `Q216_ENCOURAGE_VOTE` changes. Finally, use `mfx::logitmfx` to understand how much the probability of voting changes as `Q216_ENCOURAGE_VOTE` changes, on average. 

```{r}
# Write code here
```

### Write your interpretation below here

## Problem 8. 

Using the results of `mod_full` interpret the effect of `Q252_SAT_POL_SYSTEM`. Note if the effect is significant and explain how you came to that determination. 

Intrepret the coefficient by transforming the coefficient into an odds ratio. Then create a plot of the predicted probability of voting as `Q252_SAT_POL_SYSTEM` changes. Finally, use `mfx::logitmfx` to understand how much the probability of voting changes as `Q252_SAT_POL_SYSTEM` changes, on average. 

```{r}
# Write code here
```

### Write your interpretation below here

## Problem 9. 

Having examined the effects of each predictor which predictors do you think are the most related to voting behavior? Why?

### Write explanation below

## Problem 10. 

Create a new model named `mod_interact` that includes all of the predictors in `mod_full` and an interaction between `Q216_ENCOURAGE_VOTE` and `Q252_SAT_POL_SYSTEM` to `mod_full`.

```{r}
# Write code here

```

## Problem 11. 

In `mod_interact` we will consider `Q216_ENCOURAGE_VOTE` to be the moderator and `Q252_SAT_POL_SYSTEM` to be the predictor. Interpret the interaction, what happens to the effect of `Q252_SAT_POL_SYSTEM` at different values of `Q216_ENCOURAGE_VOTE`? 

Note if the interaction effect is significant and provide an interpretation of the odds ratio. 

To aid your interpretation, create a plot that examines how the predicted probability of voting changes as `Q252_SAT_POL_SYSTEM` at each value of `Q216_ENCOURAGE_VOTE`. 

```{r}
# Finish code below

# To get predicted probabilities for our plot, we also need to choose specific values of the other predictors. For AGE_STD we will use its mean, 0, for Q287_SES and Q282_INCOME we will use their median values, and for `Q260_SEX`, we will use F.

data_predict <- 
  tibble::tibble(
    AGE_STD = 
    Q287_SES =  
    Q288_INCOME = 
    Q260_SEX = 
    Q216_ENCOURAGE_VOTE = rep(c("Have Done", "Might Do", "Would Never Do"), 
                              each = 10), 
    Q252_SAT_POL_SYSTEM = rep(1:10, 3)
  )

data_predict <- 
  data_predict |>
  dplyr::mutate(
    pred_prob = 
  )

ggplot2::ggplot(
  data_predict, 
  ggplot2::aes(
  )
) + 
  ggplot2::facet_wrap(~) + 
  ggplot2::geom_point()
```

### Write interpretation below

## Problem 12. 

How plausible do you think the interaction above is? Does the explanation make sense to you? 

Often times we can find significant effects by chance, so it is important to have a good theoretical reason behind testing an interaction or at least very strong reasoning. 

### Write your response below

## Problem 13. 

Estimate a new model with all of the predictors in `mod_full`, but now instead of interaction between `Q252_SAT_POL_SYSTEM` and `Q216_ENCOURAGE_VOTE` add a nonlinear effect of age named `AGE_STD_SQ`. Name this model `mod_nonlinear`.  

```{r}
# Finish code below

data_analysis <- 
  data_analysis |> 
  dplyr::mutate(
    AGE_STD_SQ = 
  )

```

## Problem 14.

Interpret the nonlinear effect of age in `mod_nonlinear`, what happens to the effect of `AGE_STD` at different values of `AGE_STD`? 

Note if the nonlinear effect is significant and provide an interpretation of the odds ratio. 

To aid your interpretation, create a plot that examines how the predicted probability of voting changes as `AGE_STD` changes.

```{r}
# Finish code below

# To get predicted probabilities for our plot, we also need to choose specific values of the other predictors. For Q216_ENCOURAGE_VOTE, we will use "Might Do", for Q252_SAT_POL_SYSTEM, we will use its median value, for Q287_SES and Q282_INCOME we will use their median values, and for `Q260_SEX`, we will use F.

data_predict <- 
  tibble::tibble(
    AGE_STD = unique(data_analysis$AGE_STD), 
    Q287_SES = 
    Q288_INCOME = 
    Q260_SEX = 
    Q216_ENCOURAGE_VOTE = 
    Q252_SAT_POL_SYSTEM = 
  ) |>
  dplyr::mutate(
    AGE_STD_SQ = 
  )

data_predict <- 
  data_predict |>
  dplyr::mutate(
    pred_prob = 
  )

ggplot2::ggplot(
  data_predict, 
  ggplot2::aes(
  )
) + 
  ggplot2::geom_point()
```

### Write your response below

## Problem 15. 

Pretend you are a political scientist, look through the data frame `data_politics` and come up with three to four hypotheses about what should affect voting behavior---you can include interactions and nonlinear effects if you would like. Go through the "Purposeful Selection" steps to come to a final model and decide whether or not your hypotheses were supported.

```{r}
# Write code here
```

### Write response below here
