---
title: "Assignment 1: A Review of Statistical Concepts"
editor_options: 
  chunk_output_type: console
---

## Setup 

Load the following packages: `readr`, `tibble`, `dplyr`, `tidyr`, `ggplot2`, and `PropCIs`.

## Problem 1. 

Download `values_data.csv` from OneDrive. Create a folder named `data` and move `values_data.csv` there. Then finish the code to read in the data.

```{r}
data_values <- # Finish the code
```

## Problem 2. 

Finish the code below to recode responses to Q46 and Q47. Responses to Q46 will now be 0 and 1; 0 when the respondent is "Not very happy" or "Not at all happy" and 1 when the respondent is "Very happy" or "Rather happy." Responses to Q47 will now be 0 and 1; 0 when the respondent is "Fair", "Poor", or "Very Poor" and 1 when the respondent is "Very Good" or "Good."

```{r}
data_values <- 
  data_values |>
  dplyr::mutate(
    Q46_binary = dplyr::if_else(Q46 %in% 3:4, 0, ), # Finish code
    Q47_binary = dplyr::if_else(Q47 %in% 3:5, 0, ) # Finish code
  )
```

What proportion of respondents respondend "Very happy" or "Rather happy" to `Q46_binary`? What proportion of respondents responded `Very good` or `Good` to `Q47_binary`? 

```{r}
mean( , na.rm = TRUE) # Finish the code
mean( , na.rm = TRUE) # Finish the code
```

**Write response here**

## Problem 3. 

Use the `prop.test` function to determine if the proportion of "happy" (1) responses to `Q46_binary` is different than .50.

```{r}
total_happy <- sum( , na.rm = TRUE) # Finish the code
total_responses <- sum(!is.na(data_values$Q46_binary)) # Total non-missing responses

prop.test( , p = .50, alternative = "two.sided", correct = FALSE) # Finish the code
```

Now use `prop.test` to determine if the proportion of "happy" responses to `Q46_binary` is different than .70.

```{r}
prop.test(total_happy, total_responses, p = , alternative = "two.sided", correct = FALSE)
```

## Problem 4. 

Now finish the code below to create a new data frame that only contains responses from the USA and Canada. The column `B_COUNTRY_ALPHA` contains country codes to filter on. The US is coded `USA` and Canada is coded `CAN`. 

```{r}
data_values_us_can <- 
  data_values |>
  dplyr::filter(
     %in% c("USA", "CAN") # Finish code here
  )
```

Use the `summarize` function below to calculate the data you need to determine if the proportion of American respondents who respond "happy" to question `Q46_binary` is significantly different from the proportion of Canadian respondents who respond "happy".

```{r}
data_usa_can_summ <- 
  data_values_us_can |>
  dplyr::summarize(
    count_wellbeing_responses = sum(Q46_binary, na.rm = TRUE),
    total_responses = sum(!is.na(Q46_binary)),
    .by = # Finish code
  )
```

Using your summarized data frame,`data_us_can_summ`, finish the code below to determine if the proportion of American respondents who respond "happy" to question Q46_binary is significantly different from the proportion of Canadian respondents who respond "happy". Write up an interpretation of the `prop.test` results below. 

```{r}
prop.test(data_us_can_summ$count_happy_responses, , alternative = "two.sided", correct = FALSE)
```

**Write interpretation here**

## Problem 5. 

Using `data_values_us_can`, determine if the proportion of American respondents who respond "good" or "very good" to `Q47_binary` is significantly different from the proportion of Canadian respondents who responded "good" or "very good".

```{r}

```

## Problem 6. 

To better understand if Americans differ in their wellbeing and their health compared to Canadians calculate the absolute difference between the proportions, the relative difference, and the odds ratio.  

```{r}
data_us_can_summ <- 
  data_values_us_can |>
  dplyr::summarize(
    count_wellbeing_responses = sum(Q46_binary, na.rm = TRUE),
    total_wellbeing_responses = sum(!is.na(Q46_binary)),
    count_health_responses = sum( , na.rm = TRUE), # Finish code
    total_health_responses = sum(!is.na( )), # Finish code
    .by = # Finish code
  ) |>
  dplyr::mutate(
    prop_wellbeing = count_wellbeing_responses / total_wellbeing_responses,
    prop_health = # Finish code
  )

wellbeing_absolute_diff <- 
  data_usa_can_summ$prop_wellbeing[data_usa_can_summ$B_COUNTRY_ALPHA == "USA"] - data_usa_can_summ$prop_wellbeing[data_use_can_summ$B_COUNTRY_ALPHA == "CAN"]

wellbeing_relative_diff <- 
  data_usa_can_summ$prop_wellbeing[data_usa_can_summ$B_COUNTRY_ALPHA == "USA"] / # Finish code
  
wellbeing_odds_usa <- data_usa_can_summ$prop_wellbeing[data_usa_can_summ$B_COUNTRY_ALPHA == "USA"] / (1 - data_usa_can_summ$prop_wellbeing[data_usa_can_summ$B_COUNTRY_ALPHA == "USA"])

wellbeing_odds_can <- data_usa_can_summ$prop_wellbeing[data_usa_can_summ$B_COUNTRY_ALPHA == "CAN"] / # Finish code
  
wellbeing_odds_ratio <- wellbeing_odds_usa / wellbeing_odds_can

health_absolute_diff <- # Finish code
  
health_relative_diff <- # Finish code

health_odds_usa <- # Finish code
  
health_odds_can <- # Finish code

health_odds_ratio <- # Finish code

```

For the relative difference and odds ratios for wellbeing responses, calculate the confidence intervals using the `PropCIs` package. How do the confidence intervals support the significance test you conducted in `Problem 4`.

```{r}
count_wellbeing_usa <- data_us_can_summ$count_wellbeing_responses[data_us_can_summ$B_COUNTRY_ALPHA == "USA"]
total_wellbeing_usa <- data_us_can_summ$total_wellbeing_responses[data_us_can_summ$B_COUNTRY_ALPHA == "USA"]

count_wellbeing_can <- # Finish code
total_wellbeing_can <- # Finish code

# Relative risk confidence interval
PropCIs::riskscoreci(x1 = count_wellbeing_usa, n1 = total_wellbeing_usa,
                     x2 = , n2 = ,
                     conf.level = .95)

# Odds ratio confidence interval
PropCIs::orscoreci(x1 = count_wellbeing_usa, n1 = total_wellbeing_usa,
                     x2 = , n2 = ,
                     conf.level = .95)
```

**Write interpretation here**

## Problem 7. 

You want to understand if wellbeing as measured by the happiness question (`Q46_binary`) is related to the respondent's health as measured by the health question (`Q47`).

Finish the code below to analyze the relationship between `Q46_binary` and `Q47` by building a contingency table where `Q47` is the row variable and `Q46_binary` is the column variable. 
```{r}
xtabs( , data = data_values) # Finish the code
```

Finish the code below to calculate the joint, marginal, and conditional probabilities (probability of being happy given the respondent feels healthy) using the `prop.table` function. 

```{r}
contingency_table <- xtabs( , data = data_values) # Finish the code
contingency_table |> prop.table() |> addmargins() # Finish the code
contingency_table |> prop.table(1)
```

Using the conditional probabilities above, do you think there is a relationship between wellbeing and health? Write your response below.

**Write your response here**

## Problem 8. 

Use the `chisq.test` function to test if health and wellbeing are statistically related to one another. Using the conditional probabilities and the results of the chi-squared test, write a quick summary of the relationship between health and wellbeing.

```{r}

```

**Write interpretation here**

## Problem 9.

Use the `glm` function to estimate a simple logistic regression model where wellbeing (`Q46_binary`) is the outcome and health (`Q76`) is the predictor variable.

```{r}

```

Write a brief summary of your results. 

**Write a brief summary here**

## Problem 10.

Use the `glm` function to estimate a simple logistic regression model where wellbeing (`Q46_binary`) is the outcome and how the respondent compares to their parents standard of living  (`Q56`) is the predictor variable. Treat `Q56` as a categorical predictor. 

```{r}

```

**Write a brief summary here**